---
title: "Data Preprocessing and Visualization"
author: "Zach Schuster"
date: "December 12, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(data.table)
library(ggplot2)
library(forcats)
```


### Loading in Data

```{r load data}
# load in the data
data(algae, package = "DMwR2")

## we will use the data.table package, so turn all data frames into data tables
# here we have our training set
setDT(algae)

# we also have our testing independent variables
setDT(test.algae)

# and out testing dependent variabels, the algae samples
setDT(algae.sols)
```

***

## Data Visualization and Summarization

First, we can take a look at the structure of the data. We see that we have a few categorical variables with descriptions of the time of year and the river in which the samples were taken from. 

```{r data structure}
str(algae)

# may also be useful to look at counts of the factors
summary(algae[, sapply(algae, is.factor), with = FALSE])
```

Here we don't have too many large rivers nor do we have a lot of rivers that flow at a low speed.

***

Using the `ggplot2` package, we can create some simple visualizations of the data. We'll use the `mxPH` variable for an example.

```{r basic histogram}
ggplot(algae[!is.na(mxPH)], aes(x = mxPH)) + 
  geom_histogram(aes(y = ..density..), bins = 20)
```

A more "enriched" version of the above histogram is created below

```{r}
ggplot(algae[!is.na(mxPH)], aes(x = mxPH)) + 
  geom_histogram(aes(y = ..density..), bins = 20) +
  geom_density(color = "orange") + geom_rug() +
  labs(title = "Histogram of Maximum PH",
       x = "", y = "")

```

The data look fairly normal. We could also check this with a qqplot using the `qqPlot` function.

```{r}
car::qqPlot(algae$mxPH, main = "Normal QQ plot of Maximum PH",
            ylab = "mxPH")
```

We might be interested in checking outliers using base graphics like so:

```{r}
plot(algae$NH4, xlab = "")
abline(h = mean(algae$NH4, na.rm = TRUE), lty = 1)
abline(h = mean(algae$NH4, na.rm = TRUE) + sd(algae$NH4, na.rm = TRUE), lty = 2)
abline(h = median(algae$NH4, na.rm = TRUE), lty = 3)
identify(algae$NH4)
```

It may be interesting to look at a box plot of `a1` conditioned on river size. First, we should relevel the factors to put them in the correct order.
```{r}
algae[, `:=`(size = fct_relevel(size, c("small", "medium", "large")),
             speed = fct_relevel(speed, c("low", "medium", "high")),
             season = fct_relevel(season, c("spring", "summer", "autumn",
                                            "winter")))]

ggplot(algae, aes(x = size, y = a1)) + geom_boxplot() +
  labs(x = "River Size", y = "Algae A1")
```

Another way to create a more informative graphic than the above visualization is with a violin plot.

```{r}
ggplot(algae, aes(x = size, y = a1)) +
  geom_violin() + geom_jitter() +
  labs(x = "River Size", y = "Algae A1")
```

Finally, let's take a look at a continuous variable, `mnO2`. We will need to descretize this variable to condition on the value. 

```{r}
algae[!is.na(mnO2)][,
                    minO2 := cut(mnO2, quantile(mnO2, c(0, .25, .5, .75, 1)),
                                 include.lowest = TRUE)]

ggplot(algae[!is.na(mnO2)], aes(x = a3, y = season, color = season)) +
  geom_point() +
  facet_wrap(~ minO2) + 
  guides(color = FALSE) +
  ggtitle("")

# remove the created column
algae[, minO2 := NULL]
```

***

## Unknown Values

There are a few different ways that we can deal with missing values. I will demonstrate a few of them below. 

We can simply remove the observations with missing values as follows

```{r}
# create a new data set for the imputations
algae_rm = copy(algae)
algae_rm = algae_rm[complete.cases(algae_rm)]

# get proportion of rows remaining
nrow(algae_rm)/nrow(algae)

```

Here we removed rows using the `complete.cases` function. However, we see that 8% of our data has been removed. Ideally, we'd like to remove as little data as possible. 

***

Another way to deal with NA values is to remove rows that have a certain amount of NA values. There is a function in the `DMwR2` package, `manyNAs` to find the row indices of observations that exceed a certain threshold of NA values. We can implement our own version to deal with data.tables as well. 

```{r}
manyNA = function(data, prop, returnDT = TRUE){
  # error handling
  stopifnot(prop >= 0 && prop <= 1)
  # create boolean index
  ind = apply(data, 1, function(x){
    sum(is.na(x))/length(x) >= prop
  })
  
  # return either data.table or indices depending on returnDT arg
  if(returnDT) return(data[!ind])
  if(!returnDT) return(which(ind))
}

# create an example from the book
head(manyNA(algae, .2, returnDT = FALSE))
```

From here, we could remove the returned rows. Otherwise, the function will return a data.table with the NA heavy rows removed by default. 

***

We could consider computing with the mean or median value. We will use the median value and the `CHla` variable for example. 

```{r}
algae_rm = copy(algae)

# how many NA values are there in our variable of interest
sum(is.na(algae_rm$Chla))

# first take the median of the CHla column
CHmed = median(algae_rm$Chla, na.rm = TRUE)

# next we can impute the column of interest
algae_rm[is.na(Chla), Chla := CHmed]

#check to see how many NA values are left
sum(is.na(algae_rm$Chla))
```

We could generalize the code and use the `set` function from `data.table` to impute all numeric variables.  

***

A more advanced technique for imputation is exploring correlations between varaibles and using linear model imputations. 