---
title: "Data Preprocessing and Visualization"
author: "Zach Schuster"
date: "December 12, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# packages
library(data.table)
library(ggplot2)
```


### Loading in Data

```{r load data}
# load in the data
data(algae, package = "DMwR2")

## we will use the data.table package, so turn all data frames into data tables
# here we have our training set
setDT(algae)

# we also have our testing independent variables
setDT(test.algae)

# and out testing dependent variabels, the algae samples
setDT(algae.sols)
```

***

#* Data Visualization and Summarization

First, we can take a look at the structure of the data. We see that we have a few categorical variables with descriptions of the time of year and the river in which the samples were taken from. 

```{r data structure}
str(algae)

# may also be useful to look at counts of the factors
summary(algae[, sapply(algae, is.factor), with = FALSE])
```

Here we don't have too many large rivers nor do we have a lot of rivers that flow at a low speed.

***

Using the `ggplot2` package, we can create some simple visualizations of the data. We'll use the `mxPH` variable for an example.

```{r basic histogram}
ggplot(algae[!is.na(mxPH)], aes(x = mxPH)) + 
  geom_histogram(aes(y = ..density..), bins = 20)
```



